3D Regression Heat Map Analysis of Population Study Data

### 2015-06-25 draft.tex
added revision of Sylvia

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
$Dependent \sim Independent_1 + ... + Independent_n$.
% \begin{equation}
% Dependent \sim Independent_1 + ... + Independent_n
% \label{eq:RegressionNotation}
% \end{equation}
%%
An example of a regression formula would be $KidneyDisorder~\sim~Smoking~+~Obesity$.
The most commonly used regression operators comprise:
%%
\begin{itemize}
	\item $+,-$ inclusion/exclusion of the variable (e.g. $x\pm y$),
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion of the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
\magenta{
Due to the different meaning of operators in regression formulas, feature transformations are not available within this notation.
}
%%
The class of the target feature restricts the regression type.
%%
Different regression types are available, we focus on the following for describing linear relationships:
%%
\paragraph{Linear Regression for Continuous Target.}
%%
The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of \emph{continuous type}.
%%
\add{
%%
Linear regression models can also be described using the \emph{adjusted} $R^2$, which considers the number of dependent features.
%%
It yields lower values, when features with little entropy w.r.t. the target feature are included.
%Additional metrics describing linear regression models comprise of the \emph{adjusted} $R^2$, which also adjusts for the number of dependent features and yields lower values, when features with little entropy towards the target feature are included.
%%
The $f$-$statistic$ determines how much improvement the independent variables will add to the model compared to them being not considered at all.
%%
}

@@

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al.~\cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates as well as time plots to display and brush the data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
%%We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter.
%%
%They are therefore suitable for \emph{small} epidemiological data sets.
%%
\magenta{
Maries et al.~\cite{GRACE} propose \texttt{GRACE}, a framework for visually exploring correlations of features with spatial and non-spatial geriatric data.
%%
Sparse Partial Least Squares Regression and Tikhonov Regularization are used to produce predictor subsets and quantify correlation strengths.
%%
A multi view system linked spatial with non-spatial data and correlation coefficients are visualized using Kiviat diagrams (similar to star plots), correlations are plotted using p-values in the other views.
%%
%Dendograms are used to select different brain regions.
Instead of applying several linked views, we focus on one view of abstracted data.
}

@@

It allowed them to reproduce their knowledge by observing correlations they would expect as well as discovering new correlations.

@@

%\subsection{Abstracting Regression Results Using $R^2$}
\subsection{Abstracting Regression Results}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/heat_map}
 \caption{
 (a) Overview visualization using a 2D heat map of the formula $Z \sim X + Y$, where $Z$ assumes the feature $age$.
 %%
 The $R^2$ metrics extracted from the regression model are mapped to color saturation (a saturated color indicates a strong correlation).
 %%
 (b) Now, $Z$ is set to all features $n$ and yields $n$ 2D heat maps.
 %%
 These represent the slices in our \emph{3D Regression Heat Map}.
 %%
 The metric describing the regression model of each slice voxel is mapped on opacity in the 3D view later on, reducing the occlusion of other values.
 }
  \label{fig:Heat_Map}
\end{figure}
%%
\noindent The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%%
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a 2D heat map.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{3D Regression Heat Map} analysis we are interested in the quality-of-fit of the resulting model.
%%
This allows to infer the predictive quality of the independent features included in the model.
%%
The $R^2$ \add{, adjusted $R^2$ and AIC} value are metrics allowing for this kind of assessment (recall Sect.~\ref{sec:RegressionAnalysis}).
%%
% \add{
% %%
% Users with in-depth knowledge about regression models are given the option to render the \emph{3D Regression Heat Map} using other descriptive features presented in \ref{sec:RegressionAnalysis}.
% %%
% Therefore, these experts can investigate the cube with emphasis on specific aspects of the model.
% %%
% To avoid shutting out users unfamiliar with these metrics, the \emph{3D Regression Heat Map} is set per default to show $R^2$ values.
% }
%%
\paragraph{2D (slice) view.}
%%
Since $R^2$ is scaled between $[0, 1]$, it allows for comparison \emph{between} regression models.
%%
We can apply the same 2D heat map by translating the $R^2$ values to \emph{color saturation} (Fig.~\ref{fig:Heat_Map}a).
%%
This encodes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%
\design{
%%
Based on expert feedback on early versions of this view, the amount of features used to compare regression models was extended.
%%
Therefore, these experts can investigate the heat map with emphasis on specific aspects of the model.
%%
}
%%
\add{
%%
$Adjusted$ $R^2$ can be represented in the same way, since they share the scale.
%%
$AIC$ values have to be normalized in order to map them onto color saturation.
%%
%The resulting scale may be distort by outliers, which requires facilities for allowing the user to select scale ranges.
The resulting scale may be distorted by outliers derived from poor regression models.
%%
To tackle this problem, we provide a slider input, which maps the transfer function of the metric to color saturation based on user-selected ranges.
%%
Outliers can be cut off to emphasize ranges of interest.
%%
Small $AIC$ values indicate a good model.
%%
Hence, we inverted the transfer function mapping it to color, mapping low $AIC$ features to saturated colors.
%%
}
\design{
%%
To include users unfamiliar with these metrics, the \emph{Regression Heat Map} is set per default to show $R^2$ values.
%To avoid shutting out users unfamiliar with these metrics, the \emph{Regression Heat Map} is set per default to show $R^2$ values.
%%
}

\paragraph{3D view.}
%%
Introducing $Z$ creates a 3D heat map (Fig.~\ref{fig:Heat_Map}b).
%%
\add{The selected metric (per default set to $R^2$)} of each heat map entry (\emph{voxel}) are mapped to opacity to reduce the overlap.
%%
\magenta{Object size is not used to encode information because it would result in a cluttered view.}
%%
\design{
%%
Epidemiologists argued, that the visualization of descriptive metrics derived from different regression methods (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, we decided to map metrics of different regression methods on distinct colors (i.e., orange for linear regression and blue for logistic regression).
%%
}
%%
%The visualization of descriptive metrics derived from different 3D Regression Heat Maps (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
%Therefore, the $R^2$ results of different regression methods are encoded using different colors (i.e., orange for linear regression and blue for logistic regression).
%%
Thus, the visualization can be easily extended using other regression types.
%%
For \emph{3D Regression Heat Maps} with a fixed target feature, e.g., $Cancer \sim X + Y + Z$, no such encodings are required and the $z$ dimension can be compared directly.
%%
\magenta{
As mentioned previously, the feature reduction using the CFS algorithm potentially removes important features. %important to the domain expert.
%%
The $z$ dimension of the visualization contains \emph{all} features of the data set, allowing to assess their influence.
%%
The $x$ and $y$ dimensions are restricted to the features extracted from the CFS algorithm.
}
%%
\\\\
%%
Our goal is to create an overview visualization for a data set.
%%
We also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
The different analysis approaches require different starting points using the \emph{3D Regression Heat Map}.

@@

\paragraph{3D Regression Heat Map comparison.}
\design{
%%
Comparisons were introduced later in the project.
%%
Epidemiologists with focus on statistics pointed out, that comparing outcomes of different formulas is suitable for removing the effect of possible confounding features.
%%
}
%%
\emph{3D Regression Heat Maps} can be compared by creating difference views.
%%
One formula acts as reference.
%%
The absolute difference in the regression metric values with the second formula is calculated. %yielding the difference containing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature on the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$).

@@

\magenta{
%One limitation of the proposed method is that the regression metrics always only capture a part of the underlying model.
%%
%The analyst has to keep the aspects of the respective metric in mind to avoid false conclusions.
%%
The analysis is limited to three dynamic variables representing the \emph{3D Regression Heat Map} dimensions.
%%
Investigating more dynamic variables can be achieved by projecting the resulting high-dimensional space into a three-dimensional representation.
%%
This, however, increases the cognitive load and complexity of the analysis substantially and needs to be accompanied by techniques, which simplify this approach.
%%
Static features can be added using the formula input without increasing the complexity of the visualization.
}

>>>>

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
$Dependent \sim Independent_1 + ... + Independent_n$.
% \begin{equation}
% Dependent \sim Independent_1 + ... + Independent_n
% \label{eq:RegressionNotation}
% \end{equation}
%%
An example of a regression formula would be $KidneyDisorder~\sim~Smoking~+~Obesity$.
The most commonly used regression operators comprise:
%%
\begin{itemize}
	\item $+,-$ inclusion/exclusion of the variable (e.g. $x\pm y$),
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion of the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
\magenta{
Due to the different meaning of operators in regression formulas, feature transformations are not available within this notation.
}
%%
The class of the target feature restricts the regression type.
%%
Different regression types are available, we focus on the following for describing linear relationships:
%%
\paragraph{Linear Regression for Continuous Target.}
%%
The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of \emph{continuous type}.
%%
\add{
%%
Linear regression models can also be described using the \emph{adjusted} $R^2$, which considers the number of dependent features.
%%
It yields lower values, when features with little entropy w.r.t. the target feature are included.
%Additional metrics describing linear regression models comprise of the \emph{adjusted} $R^2$, which also adjusts for the number of dependent features and yields lower values, when features with little entropy towards the target feature are included.
%%
The $f$-$statistic$ determines how much improvement the independent variables will add to the model compared to being not considered at all.
%%
}

@@

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al.~\cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates as well as time plots to display and brush the data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
%%We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter.
%%
%They are therefore suitable for \emph{small} epidemiological data sets.
%%
\magenta{
Maries et al.~\cite{GRACE} propose \texttt{GRACE}, a framework for visually exploring correlations of features with spatial and non-spatial geriatric data.
%%
Sparse Partial Least Squares Regression and Tikhonov Regularization are used to produce predictor subsets and quantify correlation strengths.
%%
A multiple view system links spatial with non-spatial data and correlation coefficients are visualized using Kiviat diagrams (similar to star plots), correlations are plotted using p-values in the other views.
%%
%Dendograms are used to select different brain regions.
Instead of applying several linked views, we focus on one view of abstracted data.
}

@@

It allowed them to reproduce their knowledge about relationships by observing correlations they would expect as well as discovering new correlations.

@@

%\subsection{Abstracting Regression Results Using $R^2$}
\subsection{Abstracting Regression Results}
\begin{figure}[htb]
 \centering
 \includegraphics[width=1.0\linewidth]{figures/heat_map}
 \caption{
 (a) Overview visualization using a 2D heat map of the formula $Z \sim X + Y$, where $Z$ assumes the feature $age$.
 %%
 The $R^2$ metrics extracted from the regression model are mapped to color saturation (a saturated color indicates a strong correlation).
 %%
 (b) Now, $Z$ is set to all features $n$ and yields $n$ 2D heat maps.
 %%
 These represent the slices in our \emph{3D Regression Heat Map}.
 %%
 The metric describing the regression model of each slice voxel is mapped on opacity in the 3D view later on, reducing the occlusion of other values.
 }
  \label{fig:Heat_Map}
\end{figure}
%%
\noindent The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%%
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a 2D heat map.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{3D Regression Heat Map} analysis we are interested in the quality-of-fit of the resulting model.
%%
This allows to infer the predictive quality of the independent features included in the model.
%%
The $R^2$ \add{, adjusted $R^2$ and AIC} value are metrics allowing for this kind of assessment (recall Sect.~\ref{sec:RegressionAnalysis}).
%%
% \add{
% %%
% Users with in-depth knowledge about regression models are given the option to render the \emph{3D Regression Heat Map} using other descriptive features presented in \ref{sec:RegressionAnalysis}.
% %%
% Therefore, these experts can investigate the cube with emphasis on specific aspects of the model.
% %%
% To avoid shutting out users unfamiliar with these metrics, the \emph{3D Regression Heat Map} is set per default to show $R^2$ values.
% }
%%
\paragraph{2D (slice) view.}
%%
Since $R^2$ is scaled between $[0, 1]$, it allows for comparison \emph{between} regression models.
%%
We can apply the same 2D heat map by translating the $R^2$ values to \emph{color saturation} (Fig.~\ref{fig:Heat_Map}a).
%%
This encodes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%
\design{
%%
Based on expert feedback on early versions of this view, the amount of features used to compare regression models was extended.
%%
Therefore, these experts can investigate the heat map with emphasis on specific aspects of the model.
%%
}
%%
\add{
%%
\emph{Adjusted} $R^2$ can be represented in the same way, since they share the scale.
%%
$AIC$ values have to be normalized in order to map them on color saturation.
%%
%The resulting scale may be distort by outliers, which requires facilities for allowing the user to select scale ranges.
The resulting scale may be distorted by outliers derived from poor regression models.
%%
To tackle this problem, we provide a slider input, which maps the transfer function of the metric to color saturation based on user-selected ranges.
%%
Outliers can be cut off to emphasize ranges of interest.
%%
Small $AIC$ values indicate a good model.
%%
Hence, we inverted the transfer function color mapping, assigning low $AIC$ features to saturated colors.
%%
}
\design{
%%
To include users unfamiliar with these metrics, the \emph{Regression Heat Map} is set per default to show $R^2$ values.
%To avoid shutting out users unfamiliar with these metrics, the \emph{Regression Heat Map} is set per default to show $R^2$ values.
%%
}

\paragraph{3D view.}
%%
Introducing $Z$ creates a 3D heat map (Fig.~\ref{fig:Heat_Map}b).
%%
\add{The selected metric (per default set to $R^2$)} of each heat map entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
\magenta{Object size is not used to encode information because it would result in a cluttered view.}
%%
\design{
%%
Epidemiologists argued, that the visualization of descriptive metrics derived from different regression methods (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, we decided to map metrics of different regression methods on distinct colors (i.e., orange for linear regression and blue for logistic regression).
%%
}
%%
%The visualization of descriptive metrics derived from different 3D Regression Heat Maps (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
%Therefore, the $R^2$ results of different regression methods are encoded using different colors (i.e., orange for linear regression and blue for logistic regression).
%%
Thus, the visualization can be easily extended using other regression types.
%%
For \emph{3D Regression Heat Maps} with a fixed target feature, e.g., $Cancer \sim X + Y + Z$, no such encodings are required and the $z$ dimension can be compared directly.
%%
\magenta{
As mentioned previously, the feature reduction using the CFS algorithm potentially removes important features. %important to the domain expert.
%%
The $z$ dimension of the visualization contains \emph{all} features of the data set, allowing to assess their influence.
%%
The $x$ and $y$ dimensions are restricted to the features extracted from the CFS algorithm.
}
%%
\\\\
%%
Our goal is to create an overview visualization for a data set.
%%
We also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
The different analysis approaches require different starting points using the \emph{3D Regression Heat Map}.

@@

\paragraph{3D Regression Heat Map comparison.}
\design{
%%
Comparisons were introduced later in the project.
%%
Epidemiologists with focus on statistics pointed out that comparing outcomes of different formulas is suitable for removing the effect of possible confounding features.
%%
}
%%
\emph{3D Regression Heat Maps} can be compared by creating difference views.
%%
One formula acts as reference.
%%
The absolute difference in the regression metric values with the second formula is calculated. %yielding the difference containing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature on the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$).

@@

\magenta{
%One limitation of the proposed method is that the regression metrics always only capture a part of the underlying model.
%%
%The analyst has to keep the aspects of the respective metric in mind to avoid false conclusions.
%%
The analysis is limited to three dynamic variables representing the \emph{3D Regression Heat Map} dimensions.
%%
Investigating more dynamic variables can be achieved by projecting the high-dimensional space into a three-dimensional representation.
%%
This, however, increases the cognitive load and complexity of the analysis substantially and needs to be accompanied by techniques, which simplify this approach.
%%
Static features can be added using the formula input without increasing the complexity of the visualization.
}
